\textbf{Evaluation of machine learning methods for 16S rRNA gene data}

### 2019-07-08 submission/manuscript.tex
plot sizes and cowplot adjustments

Machine learning (ML) methods are effective at recognizing and
highlighting patterns in complex microbial datasets. Therefore,
researchers have started to explore the utility of ML models that use
microbiota associated biomarkers to predict human health and to
understand the microbial ecology of diseases such as liver cirrhosis,
colorectal cancer, inflammatory bowel diseases (IBD), obesity, type 2
diabetes and others (1--11). However, currently the field's use of ML
lacks clarity and consistency on which methods are used and how these
methods are implemented (12, 13). More notably, we commonly see flawed
ML practices such as using ML pipelines where there is no seperate
held-out test dataset to evaluate model performance or reporting few or
only the best outcomes of cross-validation. Even when there are seperate
testing sets to evaluate model performance, there are large differences
between cross-validation and testing performances that indicate
overfitting as well as large confidence intervals for testing
performances (4, 14--20). Moreover, there is a lack of discussion on why
a particular ML model is utilized. Recently, there is a trend towards
using more complex ML models such as random forest, extreme gradient
boosting and neural networks without a discussion on if and how much
model interpretibility is necessary for the study (11, 21--23). Black
box machine learning models require posthoc explanations to determine
the feature importances in making a prediction. These explanations can
be misleading and at times unreliable when making high-stake decisions
about someone's health (24). The lack of transparency on model selection
and interpretation as well as flawed modeling methods negatively impact
model validity and reproducibiliy. We need to strive toward better
machine learning practices by (1) implementing rigourous machine
learning pipelines and (2) selecting ML models that reflect the goal of
the study as it will inform our expectations of model accuracy,
complexity, interpretibility and computational efficiency.

To showcase a rigorous ML pipeline and to shed light on how much ML
model selection can affect modeling results, we performed an empirical
analysis comparing several different ML models using the same dataset
and the same ML pipeline. We used a previously published colorectal
cancer (CRC) study (3) which had fecal 16S rRNA gene sequences from 490
patients. CRC is a type of cancer which the human-associated human
microbiome is hypothesized to directly contribute to its development and
fecal 16S rRNA gene sequences have been used to detect CRC. We built
seven ML models using fecal 16S rRNA gene sequences to predict healthy
patients versus patients with colorectal lesions that were identified by
colonoscopy as screen relevant neoplasias (SRN). The study had 261
normal and 229 SRN samples. We established modeling pipelines for
L2-regularized logistic regression, L1 and L2 support vector machines
(SVM) with linear and radial basis function kernels, a decision tree,
random forest and XGBoost. Our ML pipeline utilized held-out test data
to evaluate predictive performance and generalizability of each ML
model. We used the area under the receiver operating characteristic
curve (AUROC) as the predictive performance metric. The median test
AUROC varied from 0.601 to 0.695. Random forest had the highest median
AUROC for detecting SRN. Despite its simplicity, the L2-regularized
logistic regression was second best in predictive performance. In terms
of computational efficiency, L1 SVM with linear kernel trained the
fastest (0.202 hours, std ± 0.028), while XGBoost took the longest
(155.104 hours, std ± 0.959). We also found that depending on how the
data is split to create a held-out test set, the AUROC values of a ML
model can vary up to 0.295 which highlighted the importance of
performing many randomized modeling runs. This study established
standards for microbiome-associated ML models and underscored the
importance of model selection.

@@

We evaluated the prediction performances of seven binary classification
models when applied to held-out test data using AUROC metric {[}Figure
2{]}. Random forest had significantly higher test AUROC values than the
other models for detecting SRNs when AUROC values were compared to the
other six by Wilcoxon rank sum test (p \textless{} 0.01). The median
AUROC of the random forest model was 0.695 (IQR 0.044). L2-regularized
logistic regression, XGBoost, L2-regularized SVM with linear and radial
basis function kernel AUROC values were not significantly different from
one another. They had median AUROC values of 0.68 (IQR 0.055), 0.679
(IQR 0.052), 0.678 (IQR 0.056) and 0.668 (IQR 0.056) respectively. L1
SVM with linear kernel and decision tree had significantly lower AUROC
values than the other ML models with median AUROC of 0.65 (IQR 0.066)
and 0.601 (IQR 0.059), respectively {[}Figure 2{]}.

@@

The ML models we built using L2-regularized logistic regression, L1 and
L2 support vector machines (SVM) with linear and radial basis function
kernels, a decision tree, random forest and XGBoost decrease in
interpretibility as they increase in complexity. We interpreted L1 and
L2 SVM with linear kernel and L2 logistic regression using the feature
weights of the trained models. We ranked the absolute weights of all the
OTUs for each data-split {[}Figure 3{]}. We calculated the median ranks
of these features over the 100 data-splits. In the three linear models,
OTUs that had the largest median ranks and drove the detection of SRNs
belonged to families \emph{Lachnospiraceae}, and \emph{Ruminococcaceae}
(OTU01239, OTU00659, OTU00742, OTU00012, OTU00015, OTU00768, OTU00822,
OTU00609), genera \emph{Gamella} (OTU00426) and genera
\emph{Peptostreptococcus} (OTU00367) {[}Figure 3{]}. Some of the OTUs
with the highest ranks were shared among the linear models. We explained
the feature importances in non-linear models using permutation
importance on the held-out test data where we randomly permuted
non-correlated features individually and groups of correlated features
together (see methods) to calculate the effect of permuted OTUs or group
of OTUs on testing AUROC. The top 5 OTUs with the largest negative
impact on testing AUROC overlapped in tree-based models {[}Figure 4{]}.
Specifically, permuting \emph{Peptostreptococcus} (OTU00367) abundances
randomly, dropped the predictive performances the most in all tree-based
methods {[}Figure S3{]}. Decision tree, random forest and XGBoost
models' predictive performance dropped from 0.6 base testing AUROC
median to 0.52, from 0.69 to 0.68 and from 0.68 to 0.65, respectively
{[}Figure 4{]}.

@@

@@

\includegraphics{Figure_1} \textbf{Figure 1. Machine learning pipeline
showing predictive model training and evaluation flowchart. } We split
the data 80\%/20\% stratified to maintain the overall label
distribution, performed five-fold cross-validation on the training data
to select the best hyperparameter setting and then using these
hyperparameters to train all of the training data. The model was
evaluated on a held-out set of data (not used in selecting the model).
Abbreviations: AUROC, area under the receiver operating characteristic
curve \newpage
\includegraphics{Figure_2.png} \textbf{Figure 2. Generalization and
classification performance of ML models using AUROC values of all cross
validation and testing performances. } The median AUROC for diagnosing
individuals with SRN using bacterial abundances was higher than chance
(depicted by horizontal line at 0.50) for all the ML models.
Discriminative performance of random forest model was higher than other
ML models. The boxplot shows quartiles at the box ends and the
statistical median as the horizontal line in the box. The whiskers show
the farthest points that are not outliers. Outliers are data points that
are not within 3/2 times the interquartile ranges. Abbreviations: SRN,
screen-relevant neoplasias; AUROC, area under the receiver operating
characteristic curve; SVM, support vector machine; XGBoost, extreme
gradient boosting \newpage
\includegraphics{Figure_3.png} \textbf{Figure 3. Interpretation of the
linear ML models.} (A) L2 logistic regression coefficients (B) L1 SVM
with linear kernel feature weights (C) L2 SVM with linear kernel feature
ranks. The feature ranks of the most important five OTUs in each
data-split are shown here. Similar OTUs had the largest impact on the
predictive performance of L2 logistic regression and L2 SVM with linear
kernel. Abbreviations: SVM, support vector machine; OTU, Operational
Taxonomic Unit. \newpage
\includegraphics{Figure_4.png} \textbf{Figure 4. Explanation of the
non-linear ML models.} (A) SVM with radial basis kernel (B) decision
tree (C) random forest (D) XGboost feaure importances were explanied
using permutation importance using held-out test set. The gray rectangle
and the dashed line show the IQR range and median of the base testing
AUROC without any permutation performed. For all the tree-based models,
a \emph{Peptostreptococcus} species (OTU00367) had the largest impact on
predictive performance of the model. Abbreviations: SVM, support vector
machine; OTU, Operational Taxonomic Unit; RBF, radial basis kernel; OTU,
Operational Taxonomic Unit. \newpage
\includegraphics{Figure_5.png} \textbf{Figure 5. Computational
efficiency of seven ML models.} The training times for training and
testing of each data-split showed the differences in computational
efficieny of the seven models. The median traintime in hours was the
highest for XGBoost and shortest for L2 logistic regression. The boxplot
shows quartiles at the box ends and the statistical median as the
horizontal line in the box. The whiskers show the farthest points that
are not outliers. Outliers are data points that are not within 3/2 times
the interquartile ranges. Abbreviations: AUROC, area under the receiver
operating characteristic curve; SVM, support vector machine; XGBoost,
extreme gradient boosting.\\
\newpage
\includegraphics{Figure_S1.png} \textbf{Figure S1. Hyperparameter
setting performances for linear models.} (A) L2 logistic regression (B)
L1 SVM with linear kernel (C) L2 SVM with linear kernel mean
cross-validation AUROC values when different hyperparameters are used in
training the model. The differences in AUROC values when hyperparameters
change show that hyperparameter tuning is a crucial step in building a
ML model.

@@

\includegraphics{Figure_S2.png} \textbf{Figure S2. Hyperparameter
setting performances for non-linear models.} (A) Decision tree (B)
Random forest (C) SVM with radial basis kernel (D) XGoost mean
cross-validation AUROC values when different hyperparameters are used in
training the model. The differences in AUROC values when hyperparameters
change show that hyperparameter tuning is a crucial step in building a
ML model. \newpage

\includegraphics{TableS1.pdf} \textbf{Table S1. ML model complexity and
hyperparameter range.} \newpage

>>>>

Machine learning (ML) methods are effective at recognizing and
highlighting patterns in complex microbial datasets. ML models learn
from existing data to predict the outcomes of new data and allow us to
infer on the reasons underlying that prediction. Therefore, researchers
have started to explore the utility of ML models that use microbiota
associated biomarkers to predict human health and to understand the
ecological basis of diseases such as liver cirrhosis, colorectal cancer,
inflammatory bowel diseases (IBD), obesity, type 2 diabetes and others
(1--11). However, currently the field's use of ML lacks clarity and
consistency on which methods are used and how these methods are
implemented (12, 13). More notably, we commonly see flawed ML practices
such as using ML pipelines where there is no seperate held-out test
dataset to evaluate model performance or reporting few or only the best
outcomes of cross-validation. Even when there are seperate testing sets
to evaluate model performance, there are large differences between
cross-validation and testing performances that indicate overfitting as
well as large confidence intervals for testing performances (4, 14--20).

Moreover, there is a lack of discussion on why a particular ML model is
utilized. Recently, there is a trend towards using more complex ML
models such as random forest, extreme gradient boosting and neural
networks without a discussion on if and how much model interpretibility
is necessary for the study (11, 21--23). Black box machine learning
models require posthoc explanations to determine the feature importances
in making a prediction. These explanations can be misleading and at
times unreliable when making high-stake decisions about someone's health
(24). The lack of transparency on model selection and interpretation as
well as flawed modeling methods negatively impact model validity and
reproducibiliy. We need to strive toward better machine learning
practices by (1) implementing rigourous machine learning pipelines and
(2) selecting ML models that reflect the goal of the study as it will
inform our expectations of model accuracy, complexity, interpretibility
and computational efficiency.

To showcase a rigorous ML pipeline and to shed light on how much ML
model selection can affect modeling results, we performed an empirical
analysis comparing several different ML models using the same dataset
and the same ML pipeline. We used a previously published colorectal
cancer (CRC) study (3) which had fecal 16S rRNA gene sequences of 490
patients. CRC is a type of cancer which the human-associated human
microbiome is hypothesized to directly contribute to its development and
fecal 16S rRNA gene sequences have been used to detect CRC. We built
seven ML models using fecal 16S rRNA gene sequences to predict healthy
patients versus patients with colorectal lesions that were identified by
colonoscopy as screen relevant neoplasias (SRN). The study had 261
normal and 229 SRN samples. We established modeling pipelines for
L2-regularized logistic regression, L1 and L2 support vector machines
(SVM) with linear and radial basis function kernels, a decision tree,
random forest and XGBoost. Our ML pipeline utilized held-out test data
to evaluate predictive performance and generalizability of each ML
model. We used the area under the receiver operating characteristic
curve (AUROC) as the predictive performance metric. The median test
AUROC varied from 0.601 to 0.695. Random forest had the highest median
AUROC for detecting SRN. Despite its simplicity, the L2-regularized
logistic regression was second best in predictive performance. In terms
of computational efficiency, L1 SVM with linear kernel trained the
fastest (0.202 hours, std ± 0.028), while XGBoost took the longest
(155.104 hours, std ± 0.959). We also found that depending on how the
data is split to create a held-out test set, the AUROC values of a ML
model can vary up to 0.295 which highlighted the importance of
performing many randomized modeling runs. This study established
standards for microbiome-associated ML models and underscored the
importance of model selection.

@@

We evaluated the predictive performances of seven binary classification
models when applied to held-out test data using AUROC metric {[}Figure
2{]}. Random forest had significantly higher test AUROC values than the
other models for detecting SRNs when AUROC values were compared to the
other six by Wilcoxon rank sum test (p \textless{} 0.01). The median
AUROC of the random forest model was 0.695 (IQR 0.044). L2-regularized
logistic regression, XGBoost, L2-regularized SVM with linear and radial
basis function kernel AUROC values were not significantly different from
one another. They had median AUROC values of 0.68 (IQR 0.055), 0.679
(IQR 0.052), 0.678 (IQR 0.056) and 0.668 (IQR 0.056) respectively. L1
SVM with linear kernel and decision tree had significantly lower AUROC
values than the other ML models with median AUROC of 0.65 (IQR 0.066)
and 0.601 (IQR 0.059), respectively {[}Figure 2{]}.

@@

The ML models we built using L2-regularized logistic regression, L1 and
L2 support vector machines (SVM) with linear and radial basis function
kernels, a decision tree, random forest and XGBoost decrease in
interpretibility as they increase in complexity. We interpreted L1 and
L2 SVM with linear kernel and L2 logistic regression using the absolute
feature weights of the trained models. We ranked the absolute weights of
all the OTUs for each data-split {[}Figure 3{]}. We calculated the
median ranks of these features over the 100 data-splits. In the three
linear models, OTUs that had the largest median ranks and drove the
detection of SRNs belonged to families \emph{Lachnospiraceae}, and
\emph{Ruminococcaceae} (OTU01239, OTU00659, OTU00742, OTU00012,
OTU00015, OTU00768, OTU00822, OTU00609), genera \emph{Gamella}
(OTU00426) and genera \emph{Peptostreptococcus} (OTU00367) {[}Figure
3{]}. Some of the OTUs with the highest ranks were shared among the
linear models.

We explained the feature importances in non-linear models, SVM with
radial basis kernel, decision tree, random forest and XGBoost, using
permutation importance on the held-out test data where we randomly
permuted non-correlated features individually and groups of correlated
features together (see methods) to calculate the effect of permuted OTUs
or group of OTUs on testing AUROC. The top 5 OTUs with the largest
negative impact on testing AUROC overlapped in tree-based models
{[}Figure 4{]}. Specifically, permuting \emph{Peptostreptococcus}
(OTU00367) abundances randomly, dropped the predictive performances the
most in all tree-based methods {[}Figure S3{]}. Decision tree, random
forest and XGBoost models' predictive performance dropped from 0.6 base
testing AUROC median to 0.52, from 0.69 to 0.68 and from 0.68 to 0.65,
respectively {[}Figure 4{]}.

We used permutation importance to interpret linear models as well to
highlight the differences between the two interpretation methods
{[}Figure S3{]}. The rankings of OTUs that discriminate between having
an SRN or not, changed between different interpretation methods.
L1-regularized SVM with linear kernel picked out some of the same OTUs
(OTU00822, OTU01239, OTU00609) as important in feature rankings
{[}Figure 3{]} and permutation importance methods {[}Figure S3{]}.
However, in general L2-regularized models had different OTUs picked out
as important in the two methods of interpretation due to collinearity in
microbial communities.

@@

Microbiome studies use ML models with a classification task to learn the
training data to assign labels, healthy or not, to new data (e.g.~new
samples), but also to learn which features are important enough to
discriminate between labels. This is a difficult task because we need to
correctly interpret our ML models. In this study we used two methods to
interpret our models. Absolute feature weights of trained models and
permutation importance. When interpreting linear models with the
differences we observed between the two interpretation methods is
possibly due to collinearity in the dataset. Collinearity is that one
OTU is highly correlated with other OTUs which means that the regression
coefficients/weights are not unique and that they are influenced by each
other. If they are influenced and not unique then it becomes harder to
interpret the model. You may want to use different models for prediction
and one for interpretation/inference, in one you remove highly
correlated OTUs to interpret and in other keep them all to predict.

@@

\includegraphics{Figure_1} \textbf{Figure 1. Machine learning pipeline
showing predictive model training and evaluation flowchart. } We split
the data 80\%/20\% stratified to maintain the overall label
distribution, performed five-fold cross-validation on the training data
to select the best hyperparameter setting and then using these
hyperparameters to train all of the training data. The model was
evaluated on a held-out set of data (not used in selecting the model).
Abbreviations: AUROC, area under the receiver operating characteristic
curve \newpage
\includegraphics{Figure_2.png}

\textbf{Figure 2. Generalization and classification performance of ML
models using AUROC values of all cross validation and testing
performances. } The median AUROC for diagnosing individuals with SRN
using bacterial abundances was higher than chance (depicted by
horizontal line at 0.50) for all the ML models. Discriminative
performance of random forest model was higher than other ML models. The
boxplot shows quartiles at the box ends and the statistical median as
the horizontal line in the box. The whiskers show the farthest points
that are not outliers. Outliers are data points that are not within 3/2
times the interquartile ranges. Abbreviations: SRN, screen-relevant
neoplasias; AUROC, area under the receiver operating characteristic
curve; SVM, support vector machine; XGBoost, extreme gradient boosting
\newpage
\includegraphics[height=15cm, width=10cm]{Figure_3.png}

\textbf{Figure 3. Interpretation of the linear ML models.} The absolute
feature weights of (A) L2 logistic regression coefficients (B) L1 SVM
with linear kernel (C) L2 SVM with linear kernel were ranked from
highest rank 1 to 100 for each data-split. The feature ranks of the
highest ranked five OTUs based on their median ranks are shown here.
Similar OTUs had the largest impact on the predictive performance of L2
logistic regression and L2 SVM with linear kernel. Abbreviations: SVM,
support vector machine; OTU, Operational Taxonomic Unit. \newpage
\includegraphics{Figure_4.png} \textbf{Figure 4. Explanation of the
non-linear ML models.} (A) SVM with radial basis kernel (B) decision
tree (C) random forest (D) XGboost feaure importances were explanied
using permutation importance using held-out test set. The gray rectangle
and the dashed line show the IQR range and median of the base testing
AUROC without any permutation performed. The colors of the boxplots
stand for the unique OTUs that are shared among the different models;
pink for OTU0008, salmon for OTU0050, yellow for OTU00367, blue for
OTU00110, green for OTU00361 and red for OTU00882. For all the
tree-based models, a \emph{Peptostreptococcus} species (OTU00367) had
the largest impact on predictive performance of the model.
Abbreviations: SVM, support vector machine; OTU, Operational Taxonomic
Unit; RBF, radial basis kernel; OTU, Operational Taxonomic Unit.
\newpage
\includegraphics{Figure_5.png} \textbf{Figure 5. Computational
efficiency of seven ML models.} The training times for of each
data-split showed the differences in computational efficieny of the
seven models. The median training time in hours was the highest for
XGBoost and shortest for L1-regularized SVM with linear kernel. The
boxplot shows quartiles at the box ends and the statistical median as
the horizontal line in the box. The whiskers show the farthest points
that are not outliers. Outliers are data points that are not within 3/2
times the interquartile ranges. Abbreviations: AUROC, area under the
receiver operating characteristic curve; SVM, support vector machine;
XGBoost, extreme gradient boosting.\\
\newpage
\includegraphics{Figure_S1.png} \textbf{Figure S1. Hyperparameter
setting performances for linear models.} (A) L2 logistic regression (B)
L1 SVM with linear kernel (C) L2 SVM with linear kernel mean
cross-validation AUROC values when different hyperparameters are used in
training the model. The differences in AUROC values when hyperparameters
change show that hyperparameter tuning is a crucial step in building a
ML model.

@@

\includegraphics{Figure_S2.png} \textbf{Figure S2. Hyperparameter
setting performances for non-linear models.} (A) Decision tree (B)
Random forest (C) SVM with radial basis kernel (D) XGoost mean
cross-validation AUROC values when different hyperparameters are used in
training the model. The differences in AUROC values when hyperparameters
change show that hyperparameter tuning is a crucial step in building a
ML model. \newpage
\includegraphics{Figure_S4.png} \textbf{Figure S3. Interpretation of the
linear ML models with permutation importance.} (A) L1-regularized SVM
with linear kernel (B) L2-regularized SVM with linear kernel and (C)
L2-regularized logistic regression were interpreted using permutation
importance using held-out test set. The gray rectangle and the dashed
line show the IQR range and median of the base testing AUROC without any
permutation performed. Abbreviations: SVM, support vector machine; OTU,
Operational Taxonomic Unit; RBF, radial basis kernel; OTU, Operational
Taxonomic Unit. \includegraphics{TableS1.pdf} \textbf{Table S1. ML model
complexity and hyperparameter range.} \newpage
